RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition

### 2020-04-15 docs/paper/paper.tex
docs: Fix some typos

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes. Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces can not be created. This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results can not be reproduced.

@@

Above version of Monte Carlo Search is greedy, because it ignores that some subtrees might be underexplored. A well-known method that explores more than greedy search, but uses less computational resources than exhaustive search, is called beam search. Instead of just one stub, $w$ distinct stubs are constructed and used, where $w$ is a hyperparameter named beam width. It is natural to call this method Monte Carlo Beam Search.

@@

Here, five rules of voice leading are imposed:
\begin{itemize}
	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
	\item Movement in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
\end{itemize}
In addition, there are three rules of harmony:
\begin{itemize}
	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
	\item Within a measure, all pitch classes must be different unless the measure is the first one or the last one;
	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
\end{itemize}

@@

In Table \ref{table:scores}, scores of a piece found with Monte Carlo Beam Search are compared with benchmarks. The piece was found with beam width set to 5 and number of trials schedule set to ($10000 \times 3$, $5000 \times 3$, $1000 \times 8$) -- this means that the second measure of the piece was chosen after 10000 random continuations and, for example, the eighth measure of the piece was chosen after 1000 random continuations. One of the benchmarks is achieved by Monte Carlo Search with 43000 random trials (i.e., total number of random trials is the same). The other benchmark relates to a piece manually created by the author in about 10 minutes. Ideally, scores should be collected during multiple independent experiments and then significance of difference should be tested, but it is not done yet.

@@

Subjective evaluation of music quality can be used for revising setup of the problem. Probably, 16 measures is too much, because long step motions are resembling descending scales.

>>>>

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes. Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces cannot be created. This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results cannot be reproduced.

@@

Above version of Monte Carlo Search is greedy, because it ignores that some subtrees might be underexplored. A well-known method that explores more than greedy search but uses less computational resources than exhaustive search, is called beam search. Instead of just one stub, $w$ distinct stubs are constructed and used, where $w$ is a hyperparameter named beam width. It is natural to call this method Monte Carlo Beam Search.

@@

Here, five rules of voice leading are imposed:
\begin{itemize}
	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
	\item Movement by step in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
\end{itemize}
In addition, there are three rules of harmony:
\begin{itemize}
	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
	\item Within a measure, all pitch classes must be different unless the measure is the first one or the last one;
	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
\end{itemize}

@@

In Table \ref{table:scores}, scores of a piece found with Monte Carlo Beam Search are compared with benchmarks. The piece was found with beam width set to 5 and number of trials schedule set to ($10000 \times 3$, $5000 \times 3$, $1000 \times 8$) -- this means that the second measure of the piece was chosen after 10000 random continuations and, for example, the eighth measure of the piece was chosen after 1000 random continuations. One of the benchmarks is achieved by Monte Carlo Search with 53000 random trials (i.e., total number of random trials is the same). The other benchmark relates to a piece manually created by the author in about 10 minutes. Ideally, scores should be collected during multiple independent experiments and then significance of difference should be tested, but it is not done yet.

@@

Subjective evaluation of music quality can be used for revising setup of the problem. Probably, 16 measures are too much, because long step motions are resembling descending scales.
